---
title: "Forest to TWS JDI Data Exploration"
format: html
editor: visual
---

# Aims

This document outlines data exploration steps of the data used to model temporal turnover in land cover cells undergoing the Forest to Transitional Woodland Shrub transition.

```{r libraries}
#| label: packages
#| echo: false
#| include: false
library(here)
source(here("scripts", "0_setup.R"))
```

```{r data preparation}
#| label: data-prep
#| include: false

# Load Turnover data
load(here("data", "derived_data",
          "all_periods_turnover_all_land_cover_chanegs_15km.rda"))

# Select only columns for Forest -> TWS model
turnover_forest_tws_15km <- all_periods_turnover_all_land_cover_chanegs_15km |>
  select(-c('2000-2006_TWS no change', '2000-2006_TWS to Forest',
            '2000-2006_Urban_no_change', '2000-2006_all_to_urban',
            '2006-2012_TWS no change', '2006-2012_TWS to Forest',
            '2006-2012_Urban_no_change', '2006-2012_all_to_urban',
            '2012-2018_TWS no change', '2012-2018_TWS to Forest',
            '2012-2018_Urban_no_change', '2012-2018_all_to_urban')) |>
  # determine which rows belong to which time period
  mutate(forest_no_change = case_when(lc_time_period == "2000-2006" ~ `2000-2006_Forest no change`,
                                      lc_time_period == "2006-2012" ~ `2006-2012_Forest no change`,
                                      lc_time_period == "2012-2018" ~ `2012-2018_Forest no change`,
                                      TRUE ~ NA_real_),
         forest_to_tws = case_when(lc_time_period == "2000-2006" ~ `2000-2006_Forest to TWS`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_Forest to TWS`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_Forest to TWS`,
                                   TRUE ~ NA_real_)) |>
  # remove columns no longer required
  select(-`2000-2006_Forest no change`, -`2006-2012_Forest no change`, 
         -`2012-2018_Forest no change`,-`2000-2006_Forest to TWS`, 
         -`2006-2012_Forest to TWS`, -`2012-2018_Forest to TWS`)

# Transform JDI values for beta regression 

# Get N
N <- nrow(turnover_forest_tws_15km)

# Calculate new JDI values
turnover_forest_tws_15km <- turnover_forest_tws_15km |>
  mutate(JDI_beta = (JDI * (N - 1) + 0.5) / N)

```

# Summary Statistics

```{r summary stats}
#| label: summary-stats
#| echo: false

# Extract summary data
summary_by_period <- turnover_forest_tws_15km |>
  group_by(lc_time_period) |>
  summarize(n_observations = n(),
            mean_JDI = mean(JDI, na.rm = TRUE),
            sd_JDI = sd(JDI, na.rm = TRUE),
            mean_forest_to_tws = mean(forest_to_tws, na.rm = TRUE),
            sd_forest_to_tws = sd(forest_to_tws, na.rm = TRUE),
            mean_forest_no_change = mean(forest_no_change, na.rm = TRUE),
            sd_forest_no_change = sd(forest_no_change, na.rm = TRUE),
            mean_delta_recorder_effort = mean(delta_recorder_effort, na.rm = TRUE),
            mean_recorder_effort = mean(recorder_effort, na.rm = TRUE))

# Create table for period summary
summary_period_table <- summary_by_period |>
  kbl(caption = "Table 1: Summary Statistics by Time Period",
      digits = 3,
      align = c("l", rep("r", ncol(summary_by_period)-1))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, background = "#E6F0FF") |>
  add_header_above(c(" " = 1, 
                     "Sample" = 1,
                     "Turnover" = 2, 
                     "Forest to TWS" = 2,
                     "Forest No Change" = 2,
                     "Recorder Effort" = 2))

# Print the table
summary_period_table

```

# Distribution of Variables

Firstly, we look at the distribution of values across each of the response variables used in our model.

```{r variables}
#| label: variables-distributions
#| echo: false

# Jaccard's Dissimilarity Index
jdi_hist <- ggplot(turnover_forest_tws_15km, aes(x = JDI)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Jaccard Dissimilarity Index",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Forest -> TWS
forest_to_tws_hist <- ggplot(turnover_forest_tws_15km, aes(x = forest_to_tws)) +
  geom_histogram(bins = 30, fill = "brown", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Forest to TWS Transition (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Forest No Change
forest_no_change_hist <- ggplot(turnover_forest_tws_15km, aes(x = forest_no_change)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Forest No Change (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Recorder effort
recorder_effort_plot <- ggplot(turnover_forest_tws_15km, aes(x = recorder_effort)) +
  geom_histogram(bins = 30, fill = "orange", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Delta recorder effort
delta_recorder_effort_plot <- ggplot(turnover_forest_tws_15km, aes(x = delta_recorder_effort)) +
  geom_histogram(bins = 30, fill = "coral", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Delta Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Combine plots
plots <- plot_grid(jdi_hist, forest_to_tws_hist, forest_no_change_hist, recorder_effort_plot, delta_recorder_effort_plot, ncol = 1, labels = c("a)", "b)", "c)", "d)", "e)"))

# Print plots
jdi_hist
forest_to_tws_hist
forest_no_change_hist
recorder_effort_plot
delta_recorder_effort_plot
```

# Relationships

## Correlation Matrix

Then, we will also check the correlation between the variables used in the analysis.

```{r variable relationships}
#| label: variables-relationships
#| echo: false

# Correlation matrix for numeric variables
numeric_vars <- turnover_forest_tws_15km |>
  select(JDI, forest_to_tws, forest_no_change, 
         intersection_size, union_size, 
         total_spp_before, total_spp_after,
         total_occ_before, total_occ_after,
         recorder_effort, delta_recorder_effort)

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Create correlation plot
cor_plot <- corrplot(cor_matrix, 
                     method = "circle", 
                     type = "upper", 
                     tl.col = "black", 
                     tl.srt = 45, 
                     addCoef.col = "black", 
                     number.cex = 0.7,
                     diag = FALSE)
```

Fortunately, we see no concerningly large correlations between variables used in the analysis.

## JDI & Forest Cells

```{r JDI values in Forest cells}
#| label: JDI-Forest
#| echo: false
#| message: false

# Scatterplot of JDI values in Forest -> TWS cells
jdi_vs_forest_to_tws <- ggplot(turnover_forest_tws_15km, 
                              aes(x = forest_to_tws, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Forest to TWS Transition (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic() +
  theme(legend.position = "none")

# Scatterplot of JDI values in Forest No Change cells
jdi_vs_forest_no_change <- ggplot(turnover_forest_tws_15km, 
                                aes(x = forest_no_change, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Forest No Change (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Display plots
# jdi_vs_forest_to_tws
# jdi_vs_forest_no_change

# Combine plots
jdi_forest_combied <- plot_grid(jdi_vs_forest_to_tws, jdi_vs_forest_no_change,
                                labels = c("a)", "b)"))

# Display combined plot
jdi_forest_combied
```

Jaccard's Dissimilarity Index appears to have a non-linear relationship with the amount of pixels of forest that are changing to TWS in 2012-2018. For the 2000-2006 period, there seems to be a rather large outlier for Forest -\> TWS (\~2000 pixels), which could bias the results of the models.

In the case of the relationship between JDI and the amount of forest that does not undergo any land cover changes, it appears that there is no relationship in the first two time periods, but a non-linear relationship in the 2012-2018 period.

## JDI & Recorder Effort

```{r}
#| label: JDI-Recorder-Effort
#| echo: false
#| message: false

# Set a global option to prevent scientific notation
options(scipen = 999)

# Scatterplots of JDI vs recorder effort 
jdi_vs_recorder_effort <- ggplot(turnover_forest_tws_15km, 
                               aes(x = recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_x_continuous(labels = label_number()) +
  scale_color_viridis_d() +
  theme_classic()+
  theme(legend.position = "none")

# Scatterplots of JDI vs delta recorder effort 
jdi_vs_delta_recorder_effort <- ggplot(turnover_forest_tws_15km, 
                                     aes(x = delta_recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Delta Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Display plots
jdi_vs_recorder_effort
jdi_vs_delta_recorder_effort

# Combine plots
jdi_recordinf_combied <- plot_grid(jdi_vs_recorder_effort, jdi_vs_delta_recorder_effort,
                                labels = c("a)", "b)"))

# Display combined plot
jdi_recordinf_combied
```

The relationship between JDI and recorder effort appears complex and differs depending on time period. In the 2000-2006, we first see a decline in JDI with increasing recording effort, followed by an increase after \~100 000 records. For the second period, we see a non-linear decline with increasing recording effort. Lastly, in the 2012-2018, the relationship does not seems to differ much from 0, but the estimation of the effect is likely impacted by the large outlier (\> 400 000 records). This appears to be something that should be accounted for.

When looking at the relationship between JDI and $\Delta Recorder_{effort}$, we generally see non-linear positive relationships between the two. In the case of the first and last periods, JDI tends to increase steeply with changes in recording effort greater than 0.5.
