---
title: "TWS to Forest JDI Data Exploration"
format: html
editor: visual
---

# Aims

This document outlines data exploration steps of the data used to model temporal turnover in land cover cells undergoing the Transitional Woodland Shrub to Forest transition.

```{r libraries}
#| label: packages
#| echo: false
#| include: false
library(here)
source(here("scripts", "0_setup.R"))
```

```{r data prep}
#| label: data-prep
#| include: false

# Load Turnover data
load(here("data", "derived_data",
          "all_periods_turnover_all_land_cover_chanegs_15km.rda"))

# Prepare data
turnover_tws_forest_15km <- all_periods_turnover_all_land_cover_chanegs_15km |>
  select(-c('2000-2006_Forest no change', '2000-2006_Forest to TWS',
            '2000-2006_Urban_no_change', '2000-2006_all_to_urban',
            '2006-2012_Forest no change', '2006-2012_Forest to TWS',
            '2006-2012_Urban_no_change', '2006-2012_all_to_urban',
            '2012-2018_Forest no change', '2012-2018_Forest to TWS',
            '2012-2018_Urban_no_change', '2012-2018_all_to_urban')) |>
  # determine which rows belong to which time period
  mutate(tws_no_change = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS no change`,
                                      lc_time_period == "2006-2012" ~ `2006-2012_TWS no change`,
                                      lc_time_period == "2012-2018" ~ `2012-2018_TWS no change`,
                                      TRUE ~ NA_real_),
         tws_to_forest = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS to Forest`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_TWS to Forest`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_TWS to Forest`,
                                   TRUE ~ NA_real_)) |>
  # remove columns no longer required
  select(-`2000-2006_TWS no change`, -`2006-2012_TWS no change`, 
         -`2012-2018_TWS no change`,-`2000-2006_TWS to Forest`, 
         -`2006-2012_TWS to Forest`, -`2012-2018_TWS to Forest`)

# Get N
N <- nrow(turnover_tws_forest_15km)

# Calculate new JDI values
turnover_tws_forest_15km <- turnover_tws_forest_15km |>
  mutate(JDI_beta = (JDI * (N - 1) + 0.5) / N)
```

# Summary Statistics

```{r summary stats}
#| label: summary-stats
#| echo: false

# Extract summary data
summary_by_period <- turnover_tws_forest_15km  |>
  group_by(lc_time_period) |>
  summarize(n_observations = n(),
            mean_JDI = mean(JDI, na.rm = TRUE),
            sd_JDI = sd(JDI, na.rm = TRUE),
            mean_tws_to_forest = mean(tws_to_forest, na.rm = TRUE),
            sd_tws_to_forest = sd(tws_to_forest, na.rm = TRUE),
            mean_tws_no_change = mean(tws_no_change, na.rm = TRUE),
            sd_tws_no_change = sd(tws_no_change, na.rm = TRUE),
            mean_delta_recorder_effort = mean(delta_recorder_effort, na.rm = TRUE),
            mean_recorder_effort = mean(recorder_effort, na.rm = TRUE))

# Create table for period summary
summary_period_table <- summary_by_period |>
  kbl(caption = "Table 1: Summary Statistics by Time Period",
      digits = 3,
      align = c("l", rep("r", ncol(summary_by_period)-1))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, background = "#E6F0FF") |>
  add_header_above(c(" " = 1, 
                     "Sample" = 1,
                     "Turnover" = 2, 
                     "TWS to Forest" = 2,
                     "TWS No Change" = 2,
                     "Recorder Effort" = 2))

# Print the table
summary_period_table
```

# Distribution of Variables

Firstly, we look at the distribution of values across each of the response variables used in our model.

```{r variables}
#| label: variables-distributions
#| echo: false

# Jaccard's Dissimilarity Index
jdi_hist <- ggplot(turnover_tws_forest_15km, aes(x = JDI)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Jaccard Dissimilarity Index",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# TWS -> Forest
tws_to_forest_hist <- ggplot(turnover_tws_forest_15km, aes(x = tws_to_forest)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Forest to TWS Transition (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Forest No Change
tws_no_change_hist <- ggplot(turnover_tws_forest_15km, aes(x = tws_no_change)) +
  geom_histogram(bins = 30, fill = "brown", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Forest No Change (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Recorder effort
recorder_effort_plot <- ggplot(turnover_tws_forest_15km, aes(x = recorder_effort)) +
  geom_histogram(bins = 30, fill = "orange", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Delta recorder effort
delta_recorder_effort_plot <- ggplot(turnover_tws_forest_15km, aes(x = delta_recorder_effort)) +
  geom_histogram(bins = 30, fill = "coral", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Delta Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))


# Print plots
jdi_hist
tws_to_forest_hist
tws_no_change_hist
recorder_effort_plot
delta_recorder_effort_plot
```

# Relationships

## Correlation Matrix

Then, we will also check the correlation between the variables used in the analysis.

```{r variable relationships}
#| label: variables-relationships
#| echo: false

# Correlation matrix for numeric variables
numeric_vars <- turnover_tws_forest_15km  |>
  select(JDI, tws_to_forest, tws_no_change, 
         intersection_size, union_size, 
         total_spp_before, total_spp_after,
         total_occ_before, total_occ_after,
         recorder_effort, delta_recorder_effort)

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Create correlation plot
cor_plot <- corrplot(cor_matrix, 
                     method = "circle", 
                     type = "upper", 
                     tl.col = "black", 
                     tl.srt = 45, 
                     addCoef.col = "black", 
                     number.cex = 0.7,
                     diag = FALSE)
```

Fortunately, we see no concerningly large correlations between variables used in the analysis.

## JDI & Transitional Woodland Shrub Cells

```{r JDI values in TWS cells}
#| label: JDI-TWS
#| echo: false
#| message: false

# Scatterplot of JDI values in Forest -> TWS cells
jdi_vs_tws_forest <- ggplot(turnover_tws_forest_15km , 
                              aes(x = tws_to_forest, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "TWS to Forest Transition (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()+
  theme(legend.position = "none")

# Scatterplot of JDI values in Forest No Change cells
jdi_vs_tws_no_change <- ggplot(turnover_tws_forest_15km, 
                                aes(x = tws_no_change, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "TWS No Change (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Combine plots
jdi_tws_combied <- plot_grid(jdi_vs_tws_forest, jdi_vs_tws_no_change,
                                labels = c("a)", "b)"))

# Display combined plot
jdi_tws_combied
```

JDI seems to have a non-linear relationship with the number of pixels that are converted from Transitional Woodland Shrub to Forest in 2000-2006 and 2012-2018. In the case of the 2006-2012 period, the relationship does not appear to be different from 0. In the case of pixels that do not change from Transitional Woodland Shrub, there is a small negative linear relationship with JDI in 2000-2006, but no relationship in 2006-2012. In the last period, the relationship is non-linear, with a steep decline before 1000 pixels, after which it increases until \~2000, followed by another decline and lastly an increase at \~5000 records.

## JDI & Recorder Effort

```{r}
#| label: JDI-Recorder-Effort
#| echo: false
#| message: false

# Scatterplots of JDI vs recorder effort 
jdi_vs_recorder_effort <- ggplot(turnover_tws_forest_15km , 
                               aes(x = recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_x_continuous(labels = label_number()) +
  scale_color_viridis_d() +
  theme_classic() +
  theme(legend.position = "none")

# Scatterplots of JDI vs delta recorder effort 
jdi_vs_delta_recorder_effort <- ggplot(turnover_tws_forest_15km , 
                                     aes(x = delta_recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Delta Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Combine plots
jdi_recordinf_combied <- plot_grid(jdi_vs_recorder_effort, jdi_vs_delta_recorder_effort,
                                labels = c("a)", "b)"))

# Display combined plot
jdi_recordinf_combied
```

The relationships between JDI and recorder effort are non-linear across all three periods of change. In the case of the relationship between JDI and $\Delta Recorder_{effort}$, JDI increases non-linearly with increasing change inn recorder effort. However, the magnitude of increase differs depending on the period considered.
