---
title: "All to Urban JDI Data Exploration"
format: html
editor: visual
---

# Aims

This document outlines data exploration steps of the data used to model temporal turnover in land cover cells undergoing the All to Urban/Artificial Surfaces transition.

```{r libraries}
#| label: packages
#| echo: false
#| include: false
library(here)
source(here("scripts", "0_setup.R"))
```

```{r data preparation}

# Load Turnover data
load(here("data", "derived_data",
          "all_periods_turnover_all_land_cover_chanegs_15km.rda"))

# Prepare data
turnover_all_urban_15km <- all_periods_turnover_all_land_cover_chanegs_15km |>
  select(-c('2000-2006_Forest no change', '2000-2006_Forest to TWS',
            '2000-2006_TWS no change', '2000-2006_TWS to Forest',
            '2006-2012_Forest no change', '2006-2012_Forest to TWS',
            '2006-2012_TWS no change', '2006-2012_TWS to Forest',
            '2012-2018_Forest no change', '2012-2018_Forest to TWS',
            '2012-2018_TWS no change', '2012-2018_TWS to Forest')) |>
  # determine which rows belong to which time period
  mutate(urban_no_change = case_when(lc_time_period == "2000-2006" ~ `2000-2006_Urban_no_change`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_Urban_no_change`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_Urban_no_change`,
                                   TRUE ~ NA_real_),
         all_to_urban = case_when(lc_time_period == "2000-2006" ~ `2000-2006_all_to_urban`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_all_to_urban`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_all_to_urban`,
                                   TRUE ~ NA_real_)) |>
  # remove columns no longer required
  select(-`2000-2006_Urban_no_change`, -`2000-2006_all_to_urban`, 
         -`2006-2012_Urban_no_change`,-`2006-2012_all_to_urban`, 
         -`2012-2018_Urban_no_change`, -`2012-2018_all_to_urban`)

# Get N
N <- nrow(turnover_all_urban_15km)

# Calculate new JDI values
turnover_all_urban_15km <- turnover_all_urban_15km |>
  mutate(JDI_beta = (JDI * (N - 1) + 0.5) / N)
```

# Summary Statistics

```{r summary stats}
#| label: summary-stats
#| echo: false

# Extract summary data
summary_by_period <- turnover_all_urban_15km  |>
  group_by(lc_time_period) |>
  summarize(n_observations = n(),
            mean_JDI = mean(JDI, na.rm = TRUE),
            sd_JDI = sd(JDI, na.rm = TRUE),
            mean_all_to_urban = mean(all_to_urban, na.rm = TRUE),
            sd_all_to_urban = sd(all_to_urban, na.rm = TRUE),
            mean_urban_no_change = mean(urban_no_change, na.rm = TRUE),
            sd_urban_no_change = sd(urban_no_change, na.rm = TRUE),
            mean_delta_recorder_effort = mean(delta_recorder_effort, na.rm = TRUE),
            mean_recorder_effort = mean(recorder_effort, na.rm = TRUE))

# Create table for period summary
summary_period_table <- summary_by_period |>
  kbl(caption = "Table 1: Summary Statistics by Time Period",
      digits = 3,
      align = c("l", rep("r", ncol(summary_by_period)-1))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, background = "#E6F0FF") |>
  add_header_above(c(" " = 1, 
                     "Sample" = 1,
                     "Turnover" = 2, 
                     "All to Urban" = 2,
                     "Urban No Change" = 2,
                     "Recorder Effort" = 2))

# Print the table
summary_period_table
```

# Distribution of Values

Firstly, we look at the distribution of values across each of the response variables used in our model.

```{r variables}
#| label: variables-distributions
#| echo: false

# Jaccard's Dissimilarity Index
jdi_hist <- ggplot(turnover_all_urban_15km, aes(x = JDI)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Jaccard Dissimilarity Index",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# All -> Urban
all_to_urban_hist <- ggplot(turnover_all_urban_15km, aes(x = tws_to_forest)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "All to Urban Transition (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Urban No Change
urban_no_change_hist <- ggplot(turnover_all_urban_15km, aes(x = tws_no_change)) +
  geom_histogram(bins = 30, fill = "brown", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Urban No Change (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Recorder effort
recorder_effort_plot <- ggplot(turnover_all_urban_15km, aes(x = recorder_effort)) +
  geom_histogram(bins = 30, fill = "orange", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Delta recorder effort
delta_recorder_effort_plot <- ggplot(turnover_all_urban_15km, aes(x = delta_recorder_effort)) +
  geom_histogram(bins = 30, fill = "coral", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Delta Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))


# Print plots
jdi_hist
all_to_urban_hist
urban_no_change_hist
recorder_effort_plot
delta_recorder_effort_plot
```

# Relationships

## Correlation Matrix

Then, we will also check the correlation between the variables used in the analysis.

```{r variable relationship}
#| label: variables-relationships
#| echo: false

# Correlation matrix for numeric variables
numeric_vars <- turnover_all_urban_15km  |>
  select(JDI, all_to_urban, urban_no_change, 
         intersection_size, union_size, 
         total_spp_before, total_spp_after,
         total_occ_before, total_occ_after,
         recorder_effort, delta_recorder_effort)

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Create correlation plot
cor_plot <- corrplot(cor_matrix, 
                     method = "circle", 
                     type = "upper", 
                     tl.col = "black", 
                     tl.srt = 45, 
                     addCoef.col = "black", 
                     number.cex = 0.7,
                     diag = FALSE)
```

Fortunately, we see no concerningly large correlations between variables used in the analysis.

# JDI and Urban Cells

```{r JDI & LC changes plots}
# Scatterplot of JDI values in Forest -> TWS cells
jdi_vs_all_urban <- ggplot(turnover_all_urban_15km , 
                            aes(x = all_to_urban, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "All to Urban Transition (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()+
  theme(legend.position = "none")

# Scatterplot of JDI values in Forest No Change cells
jdi_vs_urban_no_change <- ggplot(turnover_all_urban_15km, 
                               aes(x = urban_no_change, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Urban No Change (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Combine plots
jdi_urban_combied <- plot_grid(jdi_vs_all_urban, jdi_vs_urban_no_change,
                             labels = c("a)", "b)"))

# Display combined plot
jdi_urban_combied
```

JDI seems to have a non-linear but negative relationship with the amount of urban pixels that do not change and the number of pixels that are converted to urban areas. In the case of the 2006-2012 the relationship may be influenced by some relatively large pixel values which stand out as outliers.

# JDI and Recorder Effort

```{r}
#| label: JDI-Recorder-Effort
#| echo: false
#| message: false

# Scatterplots of JDI vs recorder effort 
jdi_vs_recorder_effort <- ggplot(turnover_all_urban_15km , 
                               aes(x = recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_x_continuous(labels = label_number()) +
  scale_color_viridis_d() +
  theme_classic() +
  theme(legend.position = "none")

# Scatterplots of JDI vs delta recorder effort 
jdi_vs_delta_recorder_effort <- ggplot(turnover_all_urban_15km , 
                                     aes(x = delta_recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Delta Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Combine plots
jdi_recordinf_combied <- plot_grid(jdi_vs_recorder_effort, jdi_vs_delta_recorder_effort,
                                labels = c("a)", "b)"))

# Display combined plot
jdi_recordinf_combied
```

The relationships between JDI and recorder effort are non-linear across all three periods of change. In the case of the relationship between JDI and $\Delta Recorder_{effort}$, JDI increases non-linearly with increasing change inn recorder effort. However, the magnitude of increase differs depending on the period considered. The relationship with recorder effort is similar to the TWS -\> Forest and the Forest -\> TWS dataframes, meaning that the relatioship is consistent regardless of the land cover change class considered.
