---
title: "TWS to Forest JDI Data Exploration"
format: html
editor: visual
number-sections: true
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
---

# Aims

This document outlines data exploration steps of the data used to model temporal turnover in land cover cells undergoing the Transitional Woodland Shrub to Forest transition.

```{r libraries}
#| label: packages
#| echo: false
#| include: false

library(here)
source(here("scripts", "0_setup.R"))
```

# All GBIF Occurrences

```{r data prep}
#| label: data-prep
#| include: false

# Load Turnover data
load(here("data", "derived_data",
          "all_periods_turnover_all_land_cover_chanegs_15km.rda"))

# Prepare data
turnover_tws_forest_15km <- all_periods_turnover_all_land_cover_chanegs_15km |>
  select(-c('2000-2006_Forest no change', '2000-2006_Forest to TWS',
            '2000-2006_Urban_no_change', '2000-2006_all_to_urban',
            '2006-2012_Forest no change', '2006-2012_Forest to TWS',
            '2006-2012_Urban_no_change', '2006-2012_all_to_urban',
            '2012-2018_Forest no change', '2012-2018_Forest to TWS',
            '2012-2018_Urban_no_change', '2012-2018_all_to_urban')) |>
  # determine which rows belong to which time period
  mutate(tws_no_change = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS no change`,
                                      lc_time_period == "2006-2012" ~ `2006-2012_TWS no change`,
                                      lc_time_period == "2012-2018" ~ `2012-2018_TWS no change`,
                                      TRUE ~ NA_real_),
         tws_to_forest = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS to Forest`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_TWS to Forest`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_TWS to Forest`,
                                   TRUE ~ NA_real_)) |>
  # remove columns no longer required
  select(-`2000-2006_TWS no change`, -`2006-2012_TWS no change`, 
         -`2012-2018_TWS no change`,-`2000-2006_TWS to Forest`, 
         -`2006-2012_TWS to Forest`, -`2012-2018_TWS to Forest`)

# Get N
N <- nrow(turnover_tws_forest_15km)

# Calculate new JDI values
turnover_tws_forest_15km <- turnover_tws_forest_15km |>
  mutate(JDI_beta = (JDI * (N - 1) + 0.5) / N)

# Prepare data for GLS: remove rows with missing x or y and categorise time periods
turnover_tws_forest_15km_coords_time <- turnover_tws_forest_15km |>
  filter(!is.na(x) & !is.na(y)) |>
  mutate(time_numeric = case_when(lc_time_period == "2000-2006" ~ 1,
                                  lc_time_period == "2006-2012" ~ 2,
                                  lc_time_period == "2012-2018" ~ 3))
# Log transform recorder effort values
turnover_tws_forest_15km_coords_time <- turnover_tws_forest_15km_coords_time |>
  mutate(log_recorder_effort = log(recorder_effort))
```

## Summary Statistics

```{r summary stats}
#| label: summary-stats
#| echo: false

# Extract summary data
summary_by_period <- turnover_tws_forest_15km  |>
  group_by(lc_time_period) |>
  summarize(n_observations = n(),
            mean_JDI = mean(JDI, na.rm = TRUE),
            sd_JDI = sd(JDI, na.rm = TRUE),
            mean_tws_to_forest = mean(tws_to_forest, na.rm = TRUE),
            sd_tws_to_forest = sd(tws_to_forest, na.rm = TRUE),
            mean_tws_no_change = mean(tws_no_change, na.rm = TRUE),
            sd_tws_no_change = sd(tws_no_change, na.rm = TRUE),
            mean_delta_recorder_effort = mean(delta_recorder_effort, na.rm = TRUE),
            mean_recorder_effort = mean(recorder_effort, na.rm = TRUE))

# Create table for period summary
summary_period_table <- summary_by_period |>
  kbl(caption = "Table 1: Summary Statistics by Time Period",
      digits = 3,
      align = c("l", rep("r", ncol(summary_by_period)-1))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, background = "#E6F0FF") |>
  add_header_above(c(" " = 1, 
                     "Sample" = 1,
                     "Turnover" = 2, 
                     "TWS to Forest" = 2,
                     "TWS No Change" = 2,
                     "Recorder Effort" = 2))

# Print the table
summary_period_table
```

Firstly, we look at the distribution of values across each of the response variables used in our model.

```{r variables}
#| label: fig-variables-distributions
#| fig-cap: "Distribution of a) Jaccard Dissimilarity Index, b) TWS to Forest transition pixel count, c) TWS No Change pixel count, d) Recorder Effort e) Change in Recorder Effort"
#| echo: false

# Jaccard's Dissimilarity Index
jdi_hist <- ggplot(turnover_tws_forest_15km, aes(x = JDI)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Jaccard Dissimilarity Index",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# TWS -> Forest
tws_to_forest_hist <- ggplot(turnover_tws_forest_15km, aes(x = tws_to_forest)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Forest to TWS Transition (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Forest No Change
tws_no_change_hist <- ggplot(turnover_tws_forest_15km, aes(x = tws_no_change)) +
  geom_histogram(bins = 30, fill = "brown", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Forest No Change (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Recorder effort
recorder_effort_plot <- ggplot(turnover_tws_forest_15km, aes(x = recorder_effort)) +
  geom_histogram(bins = 30, fill = "orange", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Delta recorder effort
delta_recorder_effort_plot <- ggplot(turnover_tws_forest_15km, aes(x = delta_recorder_effort)) +
  geom_histogram(bins = 30, fill = "coral", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Delta Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))


# Combine plots
plots <- plot_grid(jdi_hist, tws_to_forest_hist, tws_no_change_hist, recorder_effort_plot, delta_recorder_effort_plot, ncol = 2, labels = c("a)", "b)", "c)", "d)", "e)"))

# Print plots
plots
```

## Correlation Matrix

Then, we will also check the correlation between the variables used in the analysis.

```{r variable relationships}
#| label: fig-variables-relationships
#| echo: false
#| fig-cap: "Correlation matrix of variables used in models"
#| fig-width: 6
#| fig-height: 4

# Log transform recorder effort values
turnover_tws_forest_15km <- turnover_tws_forest_15km |>
  mutate(log_recorder_effort = log(recorder_effort))

# Correlation matrix for numeric variables
numeric_vars <- turnover_tws_forest_15km  |>
  select(JDI, tws_to_forest, tws_no_change, 
         intersection_size, union_size, 
         total_spp_before, total_spp_after,
         total_occ_before, total_occ_after,
         recorder_effort, delta_recorder_effort,
         log_recorder_effort)

# Calculate correlation matrix
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Create correlation plot
cor_plot <- corrplot(cor_matrix, 
                     method = "circle", 
                     type = "upper", 
                     tl.col = "black", 
                     tl.srt = 45, 
                     addCoef.col = "black", 
                     number.cex = 0.7,
                     diag = FALSE)
```

Fortunately, we see no concerningly large correlations between variables used in the analysis.

## JDI & Transitional Woodland Shrub Cells

```{r JDI values in TWS cells}
#| label: JDI-TWS-cells
#| echo: false
#| message: false
#| fig-cap: "Relationship between calculated JDI and amount of TWS to Forest and TWS No Change"
#| fig-width: 6
#| fig-height: 4

# Scatterplot of JDI values in Forest -> TWS cells
jdi_vs_tws_forest <- ggplot(turnover_tws_forest_15km , 
                              aes(x = tws_to_forest, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "TWS to Forest Transition (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()+
  theme(legend.position = "none")

# Scatterplot of JDI values in Forest No Change cells
jdi_vs_tws_no_change <- ggplot(turnover_tws_forest_15km, 
                                aes(x = tws_no_change, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "TWS No Change (pixel count)",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Combine plots
jdi_tws_combied <- plot_grid(jdi_vs_tws_forest, jdi_vs_tws_no_change,
                                labels = c("a)", "b)"))

# Display combined plot
jdi_tws_combied
```

JDI seems to have a non-linear relationship with the number of pixels that are converted from Transitional Woodland Shrub to Forest in 2000-2006 and 2012-2018. In the case of the 2006-2012 period, the relationship does not appear to be different from 0. In the case of pixels that do not change from Transitional Woodland Shrub, there is a small negative linear relationship with JDI in 2000-2006, but no relationship in 2006-2012. In the last period, the relationship is non-linear, with a steep decline before 1000 pixels, after which it increases until \~2000, followed by another decline and lastly an increase at \~5000 records.

## JDI & Recorder Effort

```{r JDI recorder effort}
#| label: JDI-recorder-effort
#| echo: false
#| message: false
#| fig-cap: "Relationship between calculated JDI and Recorder Effort"
#| fig-width: 6
#| fig-height: 4

# Scatterplots of JDI vs recorder effort 
jdi_vs_recorder_effort <- ggplot(turnover_tws_forest_15km , 
                               aes(x = recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_x_continuous(labels = label_number()) +
  scale_color_viridis_d() +
  theme_classic() +
  theme(legend.position = "none")

# Scatterplots of JDI vs delta recorder effort 
jdi_vs_delta_recorder_effort <- ggplot(turnover_tws_forest_15km , 
                                     aes(x = delta_recorder_effort, y = JDI, color = lc_time_period)) +
  geom_point(alpha = 0.5) +
  geom_smooth() +
  labs(x = "Delta Recorder Effort",
       y = "Jaccard Dissimilarity Index",
       color = "Time Period") +
  scale_color_viridis_d() +
  theme_classic()

# Combine plots
jdi_recordinf_combied <- plot_grid(jdi_vs_recorder_effort, jdi_vs_delta_recorder_effort,
                                labels = c("a)", "b)"))

# Display combined plot
jdi_recordinf_combied
```

The relationships between JDI and recorder effort are non-linear across all three periods of change. In the case of the relationship between JDI and $\Delta Recorder_{effort}$, JDI increases non-linearly with increasing change inn recorder effort. However, the magnitude of increase differs depending on the period considered.

## Modeling

### Trial 1: $\beta$ regression

To start off, we ran the following $\beta$ regression:

```{r beta regression}
#| label: tbl-beta-regression
#| echo: false
#| tbl-cap: "Beta Regression Results TWS-> Forest Land Cover Change"

# Load model
load(here("data", "models", "exploratory",
          "TWSF_turnover_model1_beta_regression.RData"))

# Extract residuals from model
model2.1_residuals <- residuals(TWSF_turnover_model1_beta_regression)

# Add residuals to df
turnover_tws_forest_15km$residuals <- model2.1_residuals

# Get summary
model_summary <- tidy(TWSF_turnover_model1_beta_regression,
                      conf.int = TRUE)

# Display summary
knitr::kable(model_summary, digits = 4)
```

```{r beta regression model fit}
#| label: tbl-beta-regression-model-fit
#| echo: false
#| warning: false
#| tbl-cap: "Beta Regression Results Model Fit Statistics"

# Dispaly model fit statistics
model_glance <- glance(TWSF_turnover_model1_beta_regression)
knitr::kable(model_glance, digits = 4)
```

### Check for Spatial Autocorrelation

```{r Morans I table check for spatial autocorrelation}
#| label: tbl-morans-i
#| echo: false
#| tbl-cap: "Moran's I Test for Spatial Autocorrelation of Model Residuals"
#| fig-width: 12
#| fig-height: 8

# Load TWS -> Forest raster
clc_status_15km_tws_forest_masked <- rast(here("data", "derived_data", 
                                               "clc_status_15km_tws_forest_masked.tif"))

# Create reference grid
reference_grid <- clc_status_15km_tws_forest_masked[[1]]

# Convert turnover df to sf object
turnover_sf <- st_as_sf(turnover_tws_forest_15km,
                        coords = c("x", "y"),
                        crs = st_crs(reference_grid))

# Create a neighbour list using k-nearest neighbours (k = 5)
coords_matrix <- st_coordinates(turnover_sf)
neighbours <- knn2nb(knearneigh(coords_matrix, k = 5))

# Convert to spatial weights matrix
weights <- nb2listw(neighbours, style = "W")

# Compute Moran's I test for spatial autocorrelation
moran_test <- moran.test(turnover_sf$residuals, listw = weights)

# Print results
moran_results <- data.frame(Statistic = c("Moran I", "Expectation", "Variance", "Standard deviate", "P-value"),
                            Value = c(moran_test$estimate[1], 
                                      moran_test$estimate[2], 
                                      moran_test$estimate[3],
                                      moran_test$statistic,
                                      moran_test$p.value))
# Display output
knitr::kable(moran_results, digits = 6,
             col.names = c("Test Statistic", "Value"))

# Create Moran scatterplot to visualise the relationship
# moran_plot <- moran.plot(turnover_sf$residuals, listw = weights,
#                          xlab = "Model Residuals", 
#                          ylab = "Spatially Lagged Residuals")
```

```{r Morans I figure}
#| label: fig-morans-i
#| echo: false
#| fig-cap: "Distribution of a) Local Moran's I, b,c) Significant Spatial Autocorrelation Values"
#| fig-width: 12
#| fig-height: 8

# Calculate local Moran's I wiht zero.policy
local_moran <- localmoran(turnover_sf$residuals, weights, zero.policy = TRUE)

# Add local Moran's I statistics to the spatial dataframe
turnover_sf$local_moran_i <- local_moran[, 1] # I statistics
turnover_sf$local_moran_p <- local_moran[, 5] # p-value

# Check for missing values in Moran's I column
# missing_values <- sum(is.na(turnover_sf$local_moran_i))
# cat("Number of cells with missing Local Moran's I values:", missing_values, "\n") #0

# Filter out NA for mapping
turnover_sf_clean <- turnover_sf |>
  filter(!is.na(local_moran_i))


# Extract coordinates
coords <- st_coordinates(turnover_sf)
turnover_df <- cbind(st_drop_geometry(turnover_sf), coords)

# Plot map
plot1 <- ggplot(turnover_df |> 
                  filter(!is.na(local_moran_i)), 
                aes(x = X, y = Y, fill = local_moran_i)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis", name = "Local Moran's I") +
  theme_classic() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank())

# Save map
# ggsave(here("figures", "SupplementaryFigure5a_Local_Moran_I_TWS_Forest_15km.png"),
#        plot1, width = 10, height = 8)


# Plot map
plot2 <- ggplot(turnover_df |> 
                  filter(!is.na(local_moran_i)), 
                aes(x = X, y = Y, fill = local_moran_p < 0.01)) +
  geom_tile() +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "grey"),
                    name = "Significant",
                    labels = c("FALSE" = "Not significant", "TRUE" = "p < 0.001")) +
  theme_classic() +
  theme(panel.grid = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank(),
          axis.line = element_blank(),
          axis.ticks = element_blank())

# Save figure 
# ggsave(here("figures", "SupplementaryFigure5b_Local_Moran_I_Significance_TWS_Forest_15km.png"),
#        plot1, width = 10, height = 8)

# Create categorical variable that combines significance and residual sign
turnover_df <- turnover_df |>
  mutate(sig_residual_type = case_when(local_moran_p < 0.01 & residuals > 0 ~ "Significant Positive",
                                       local_moran_p < 0.01 & residuals < 0 ~ "Significant Negative",
                                       local_moran_p >= 0.01 ~ "Not Significant"))

# Plot map of significant clusters with positive/negative residuals
plot3 <- ggplot(turnover_df |>
                  filter(!is.na(local_moran_i)), 
                aes(x = X, y = Y, fill = sig_residual_type)) +
  geom_tile() +
  scale_fill_manual(values = c("Significant Positive" = "pink", 
                               "Significant Negative" = "lightblue",
                               "Not Significant" = "grey"),
                    name = "Residual Type") +
  theme_classic() +
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank())

# Combine plots
tws_forest_15km_local_morans_I <- plot_grid(plot1, plot2, plot3,
                                            labels = c("a)", "b)", "c)"),
                                            nrow = 2, align = "h")

# Display plot
tws_forest_15km_local_morans_I

# Save to file
ggsave(here("figures", "SupplementaryFigure5_Local_Moran_I_TWS_Forest_15km.png"),
       tws_forest_15km_local_morans_I, width = 18, height = 8)
```

Moran's I test for spatial autocorrelation revealed that there is significant spatial autocorrelation. Therefore, this needs to be accounted for to improve model predictions. Given the testing we have done previously, for the Forest -\> TWS models, we will no longer attempt the GLMM, ordered $\beta$ or the GLS with raw data. We will go straight to trying the GLS with logged recorder effort.

## (Final) Trial 2: GLS with log recorder effort

```{r gls log correlation}
#| label: gls-logged-corr-structure
#| echo: false
#| message: false

# Load model
load(here("data", "models", "exploratory", "TWSF_turnover_model2_gls_log.RData"))

# Extract correlation structure parameters
print(TWSF_turnover_model2_gls_log$modelStruct$corStruct)

# Get range parameter
range_param <- coef(TWSF_turnover_model2_gls_log$modelStruct$corStruct, unconstrained = FALSE)
```

```{r gls log model diagnostic}
#| label: fig-gls-logged-values
#| echo: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Diagnostic plots for GLS model 2"

# Extract residuals
residuals_gls <- residuals(TWSF_turnover_model2_gls_log, type = "normalized")

# Get the model frame to ensure we're using the exact same data
model_data <- getData(TWSF_turnover_model2_gls_log)

# Basic residual plots
par(mfrow = c(2, 2))

# Residuals vs fitted
plot(fitted(TWSF_turnover_model2_gls_log), residuals_gls,
     xlab = "Fitted Values", ylab = "Normalized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

# QQ plot
qqnorm(residuals_gls, main = "Normal Q-Q Plot")
qqline(residuals_gls, col = "red")

# Residuals vs predictors
plot(model_data$tws_to_forest, residuals_gls,
     xlab = "Forest to TWS", ylab = "Normalized Residuals",
     main = "Residuals vs TWS to Forest")
abline(h = 0, col = "red", lty = 2)

plot(model_data$log_recorder_effort, residuals_gls,
     xlab = "Recorder Effort", ylab = "Normalized Residuals",
     main = "Residuals vs Recorder Effort")
abline(h = 0, col = "red", lty = 2)
```

The diagnostic plots for this model show that:

1.  Residuals vs Fitted:
    -   Most residuals are centered around 0
    -   Spread is roughly constant across the fitted values
2.  QQ Plot:
    -   There is some deviation from the red line, especially at the bottom of the tail, suggesting that residuals are not perfectly normal
    -   Our sample size is relatively large, and this is somewhat common with ecological data
3.  Residuals vs. Forest -\> TWS
    -   Data concentration at 0, which makes sense given that most of the cells will have no TWS to Forest
    -   But relatively even spread
4.  Residuals vs. Recorder Effort
    -   Residuals are evenly spread across the range of log recorder effort
    -   Residuals are (mostly) centered around zero and there is no obvious pattern observed

This model looks good to use! Therefore, it will be the one used in the manuscript.

```{r gls summary}
#| label: tbl-gls-logged-values-summary
#| echo: false
#| tbl-cap: "GLS Model Coefficients TWS → Forest Effects on Biodiversity Turnover"

# Extract model summary
model_summary <- summary(TWSF_turnover_model2_gls_log)
coef_table <- model_summary$tTable

# Create coefficient table without confidence intervals initially
model_coefs <- data.frame(Variable = rownames(coef_table),
                          Estimate = round(coef_table[, "Value"], 4),
                          `Std Error` = round(coef_table[, "Std.Error"], 4),
                          `t value` = round(coef_table[, "t-value"], 3),
                          `p value` = coef_table[, "p-value"],
                          check.names = FALSE)

# Add significance stars
model_coefs$Significance <- case_when(model_coefs$`p value` < 0.001 ~ "***",
                                      model_coefs$`p value` < 0.01 ~ "**", 
                                      model_coefs$`p value` < 0.05 ~ "*",
                                      model_coefs$`p value` < 0.1 ~ ".",
                                      TRUE ~ "")

# Display using kable
knitr::kable(model_coefs, digits = c(0, 4, 4, 3, 6, 0),
             format.args = list(scientific = FALSE))
```

```{r gls2 model fit statistics}
#| label: tbl-gls-model-fit
#| echo: false
#| tbl-cap: "GLS Model Fit Statistics"


# Model information
cat("Model Specifications:\n")
cat("- Formula:", deparse(TWSF_turnover_model2_gls_log$call$model), "\n")
cat("- Correlation Structure: Exponential spatial correlation\n") 
cat("- Method: REML\n")

# Try to get sample size safely
n_obs <- tryCatch({model_summary$dims$N
}, error = function(e) {
  nobs(TWSF_turnover_model2_gls_log)
})

cat("- Number of observations:", n_obs, "\n")

# Model fit statistics
fit_stats <- data.frame(Statistic = c("AIC", "BIC", "Log-likelihood", "Residual Std Error"),
                        Value = c(round(AIC(TWSF_turnover_model2_gls_log), 2),
                                  round(BIC(TWSF_turnover_model2_gls_log), 2), 
                                  round(as.numeric(logLik(TWSF_turnover_model2_gls_log)), 2),
                                  round(model_summary$sigma, 4)))
knitr::kable(fit_stats, col.names = c("Statistic", "Value"))
```

```{r gls model output plot}
#| label: fig-gls-coefficient-plot
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Coefficient Estimates from GLS"

# Get summary of the model
model_summary <- summary(TWSF_turnover_model2_gls_log)

# Create dataframe of coeficients
TWSF_turnover_model2_gls_log_coef_df <- data.frame(term = names(model_summary$tTable[, "Value"]),
                                                   estimate = model_summary$tTable[, "Value"],
                                                   std.error = model_summary$tTable[, "Std.Error"],
                                                   statistic = model_summary$tTable[, "t-value"],
                                                   p.value = model_summary$tTable[, "p-value"])

# Remove the intercept
TWSF_turnover_model2_gls_log_coef_df_no_intercept <- TWSF_turnover_model2_gls_log_coef_df[TWSF_turnover_model2_gls_log_coef_df$term != "(Intercept)", ]

# Create coefficient plot
figure7_a <- ggplot(TWSF_turnover_model2_gls_log_coef_df_no_intercept, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = estimate - 1.96 * std.error,
                     xmax = estimate + 1.96 * std.error)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_y_discrete(labels = c("tws_to_forest" = "TWS to Forest",
                              "tws_no_change" = "TWS No Change", 
                              "delta_recorder_effort" = "ΔRecorder Effort",
                              "log_recorder_effort" = "log Recorder Effort",
                              "lc_time_period2006-2012" = "2006-2012 Time Period",
                              "lc_time_period2012-2018" = "2012-2018 Time Period"),
                   limits = c("lc_time_period2012-2018",
                              "lc_time_period2006-2012", 
                              "log_recorder_effort",
                              "delta_recorder_effort",
                              "tws_no_change",
                              "tws_to_forest")) +
  labs(x = "Estimate ± 95% CI", y = NULL) +
  theme_classic()

# Display plot
figure7_a
```

# Vascular Plant Occurrences

```{r prepare vascular plant data}
#| label: plant-data-prep
#| include: false

# Load Turnover data
load(here("data", "derived_data", 
          "vascular_plants_all_periods_turnover_all_land_cover_chanegs_15km.rda"))

# Select only TWS -> Forest columns
plants_turnover_tws_forest_15km <- vascular_plants_all_periods_turnover_all_land_cover_chanegs_15km |>
  select(-c('2000-2006_Forest no change', '2000-2006_Forest to TWS',
            '2000-2006_Urban_no_change', '2000-2006_all_to_urban',
            '2006-2012_Forest no change', '2006-2012_Forest to TWS',
            '2006-2012_Urban_no_change', '2006-2012_all_to_urban',
            '2012-2018_Forest no change', '2012-2018_Forest to TWS',
            '2012-2018_Urban_no_change', '2012-2018_all_to_urban')) |>
  # determine which rows belong to which time period
  mutate(tws_no_change = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS no change`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_TWS no change`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_TWS no change`,
                                   TRUE ~ NA_real_),
         tws_to_forest = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS to Forest`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_TWS to Forest`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_TWS to Forest`,
                                   TRUE ~ NA_real_)) |>
  # remove columns no longer required
  select(-`2000-2006_TWS no change`, -`2006-2012_TWS no change`, 
         -`2012-2018_TWS no change`,-`2000-2006_TWS to Forest`, 
         -`2006-2012_TWS to Forest`, -`2012-2018_TWS to Forest`)

# Remove rows that might have NA for x or y and categorise time periods for GLS
plants_turnover_tws_forest_15km_coords_time <- plants_turnover_tws_forest_15km |>
  filter(!is.na(x) & !is.na(y)) |>
  mutate(time_numeric = case_when(lc_time_period == "2000-2006" ~ 1,
                                  lc_time_period == "2006-2012" ~ 2,
                                  lc_time_period == "2012-2018" ~ 3),
         log_recorder_effort = log(recorder_effort))

```

## Summary Statistics

```{r plant-summary-stats}
#| label: tbl-summary-stats-plants
#| echo: false
#| tbl-cap: "Summary Statistics of Jaccard Dissimilarity Index (JDI) by Time Period for Vascular Plant Records"

# Extract summary data
plant_summary_by_period <- plants_turnover_tws_forest_15km |>
  group_by(lc_time_period) |>
  summarize(n_observations = n(),
            mean_JDI = mean(JDI, na.rm = TRUE),
            sd_JDI = sd(JDI, na.rm = TRUE),
            mean_tws_to_forest = mean(tws_to_forest, na.rm = TRUE),
            sd_tws_to_forest = sd(tws_to_forest, na.rm = TRUE),
            mean_tws_no_change = mean(tws_no_change, na.rm = TRUE),
            sd_tws_no_change = sd(tws_no_change, na.rm = TRUE),
            mean_delta_recorder_effort = mean(delta_recorder_effort, na.rm = TRUE),
            mean_recorder_effort = mean(recorder_effort, na.rm = TRUE))

# Create table for period summary
plant_summary_by_period <- plant_summary_by_period |>
  kbl(digits = 3,
      align = c("l", rep("r", ncol(plant_summary_by_period)-1))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, background = "#E6F0FF") |>
  add_header_above(c(" " = 1, 
                     "Sample" = 1,
                     "Turnover" = 2, 
                     "TWS to Forest" = 2,
                     "TWS No Change" = 2,
                     "Recorder Effort" = 2))

# Print the table
plant_summary_by_period
```

## Distribution of Values

```{r plant-variables}
#| label: fig-variables-distributions-plant
#| fig-cap: "Distribution of a) Jaccard Dissimilarity Index, b) TWS to Forest transition pixel count, c) TWS No Change pixel count, d) Recorder Effort e) Change in Recorder Effort in Subser with Only Vascular Plant Records"
#| echo: false

# Jaccard's Dissimilarity Index
plant_jdi_hist <- ggplot(plants_turnover_tws_forest_15km, aes(x = JDI)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Jaccard Dissimilarity Index",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# TWS -> Forest
plant_tws_to_forest_hist <- ggplot(plants_turnover_tws_forest_15km, aes(x = tws_to_forest)) +
  geom_histogram(bins = 30, fill = "brown", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "TWS to Forest Transition (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# TWS No Change
plant_tws_no_change_hist <- ggplot(plants_turnover_tws_forest_15km, aes(x = tws_no_change)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "TWS No Change (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Recorder effort
plant_recorder_effort_plot <- ggplot(plants_turnover_tws_forest_15km, aes(x = recorder_effort)) +
  geom_histogram(bins = 30, fill = "orange", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Delta recorder effort
plant_delta_recorder_effort_plot <- ggplot(plants_turnover_tws_forest_15km, aes(x = delta_recorder_effort)) +
  geom_histogram(bins = 30, fill = "coral", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Delta Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Combine plots
plant_plots <- plot_grid(plant_jdi_hist, plant_tws_to_forest_hist, plant_tws_no_change_hist, plant_recorder_effort_plot, plant_delta_recorder_effort_plot, ncol = 2, labels = c("a)", "b)", "c)", "d)", "e)"))

# Print plots
plant_plots
```

## Correlation Matrix

```{r plant correlation matrix}
#| label: fig-variables-relationships-plants
#| echo: false
#| fig-cap: "Correlation matrix of variables used in the vascular plant models"
#| fig-width: 6
#| fig-height: 4

# Correlation matrix for numeric variables
plant_numeric_vars <- plants_turnover_tws_forest_15km_coords_time |>
  select(JDI, tws_to_forest, tws_no_change, 
         intersection_size, union_size, 
         total_spp_before, total_spp_after,
         total_occ_before, total_occ_after,
         recorder_effort, delta_recorder_effort,
         log_recorder_effort)

# Calculate correlation matrix
plant_cor_matrix <- cor(plant_numeric_vars, use = "pairwise.complete.obs")

# Create correlation plot
plant_cor_plot <- corrplot(plant_cor_matrix, 
                     method = "circle", 
                     type = "upper", 
                     tl.col = "black", 
                     tl.srt = 45, 
                     addCoef.col = "black", 
                     number.cex = 0.7,
                     diag = FALSE)
```

## Modeling

### (Final) Trial 1: GLS with log recorder effort

```{r gls plants log correlation}
#| label: gls-plants-logged-corr-structure
#| echo: false
#| message: false

# Load model
load(here("data", "models", "exploratory", "plants_TWSF_model1_gls.RData"))

# Extract correlation structure parameters
print(plants_TWSF_model1_gls$modelStruct$corStruct)

# Get range parameter
range_param <- coef(plants_TWSF_model1_gls$modelStruct$corStruct, unconstrained = FALSE)
```

```{r gls plants log model diagnostic}
#| label: fig-plants-gls-logged-values
#| echo: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Diagnostic plots for GLS Plants model 1"

# Extract residuals
residuals_gls <- residuals(plants_TWSF_model1_gls, type = "normalized")

# Get the model frame to ensure we're using the exact same data
model_data <- getData(plants_TWSF_model1_gls)

# Basic residual plots
par(mfrow = c(2, 2))

# Residuals vs fitted
plot(fitted(plants_TWSF_model1_gls), residuals_gls,
     xlab = "Fitted Values", ylab = "Normalized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

# QQ plot
qqnorm(residuals_gls, main = "Normal Q-Q Plot")
qqline(residuals_gls, col = "red")

# Residuals vs predictors
plot(model_data$tws_to_forest, residuals_gls,
     xlab = "TWS to Forest", ylab = "Normalized Residuals",
     main = "Residuals vs TWS to Forest")
abline(h = 0, col = "red", lty = 2)

plot(model_data$log_recorder_effort, residuals_gls,
     xlab = "Recorder Effort", ylab = "Normalized Residuals",
     main = "Residuals vs Recorder Effort")
abline(h = 0, col = "red", lty = 2)
```

The diagnostic plots for this model show that:

1.  Residuals vs Fitted:
    -   Most residuals are centered around 0
    -   Spread is roughly constant across the fitted values but there is some funneling
2.  QQ Plot:
    -   There is some deviation from the red line at both ends suggesting that residuals are not perfectly normal
    -   Our sample size is relatively large, and this is somewhat common with ecological data
3.  Residuals vs. Forest -\> TWS
    -   Data concentration at 0, which makes sense given that most of the cells will have no TWS to Forest
    -   But relatively even spread and no strong pattern
4.  Residuals vs. Recorder Effort
    -   Residuals are evenly spread across the range of log recorder effort
    -   Residuals are (mostly) centered around zero and there is no obvious pattern observed

This model looks ok(ish) to use! Therefore, it will be the one used in the manuscript.

```{r gls plants summary}
#| label: tbl-gls-logged-values-summary
#| echo: false
#| tbl-cap: "GLS Model Coefficients TWS → Forest Effects on Vascular Plant Turnover"

# Extract model summary
model_summary <- summary(plants_TWSF_model1_gls)
coef_table <- model_summary$tTable

# Create coefficient table without confidence intervals initially
model_coefs <- data.frame(Variable = rownames(coef_table),
                          Estimate = round(coef_table[, "Value"], 4),
                          `Std Error` = round(coef_table[, "Std.Error"], 4),
                          `t value` = round(coef_table[, "t-value"], 3),
                          `p value` = coef_table[, "p-value"],
                          check.names = FALSE)

# Add significance stars
model_coefs$Significance <- case_when(model_coefs$`p value` < 0.001 ~ "***",
                                      model_coefs$`p value` < 0.01 ~ "**", 
                                      model_coefs$`p value` < 0.05 ~ "*",
                                      model_coefs$`p value` < 0.1 ~ ".",
                                      TRUE ~ "")

# Display using kable
knitr::kable(model_coefs, digits = c(0, 4, 4, 3, 6, 0),
             format.args = list(scientific = FALSE))
```

```{r gls2 plant model fit statistics}
#| label: tbl-gls-model-fit
#| echo: false
#| tbl-cap: "GLS Plant Model Fit Statistics"


# Model information
cat("Model Specifications:\n")
cat("- Formula:", deparse(plants_TWSF_model1_gls$call$model), "\n")
cat("- Correlation Structure: Exponential spatial correlation\n") 
cat("- Method: REML\n")

# Try to get sample size safely
n_obs <- tryCatch({model_summary$dims$N
}, error = function(e) {
  nobs(plants_TWSF_model1_gls)
})

cat("- Number of observations:", n_obs, "\n")

# Model fit statistics
fit_stats <- data.frame(Statistic = c("AIC", "BIC", "Log-likelihood", "Residual Std Error"),
                        Value = c(round(AIC(plants_TWSF_model1_gls), 2),
                                  round(BIC(plants_TWSF_model1_gls), 2), 
                                  round(as.numeric(logLik(plants_TWSF_model1_gls)), 2),
                                  round(model_summary$sigma, 4)))
knitr::kable(fit_stats, col.names = c("Statistic", "Value"))
```

```{r gls plant model output plot}
#| label: fig-gls-plant-coefficient-plot
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Coefficient Estimates from Plant GLS"

# Get summary of the model
model_summary <- summary(plants_TWSF_model1_gls)

# Create dataframe of coeficients
plants_TWSF_model1_gls_coef_df <- data.frame(term = names(model_summary$tTable[, "Value"]),
                                                   estimate = model_summary$tTable[, "Value"],
                                                   std.error = model_summary$tTable[, "Std.Error"],
                                                   statistic = model_summary$tTable[, "t-value"],
                                                   p.value = model_summary$tTable[, "p-value"])

# Remove the intercept
plants_TWSF_model1_gls_coef_df_no_intercept <- plants_TWSF_model1_gls_coef_df[plants_TWSF_model1_gls_coef_df$term != "(Intercept)", ]

# Create coefficient plot
figure7_b <- ggplot(plants_TWSF_model1_gls_coef_df_no_intercept, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = estimate - 1.96 * std.error,
                     xmax = estimate + 1.96 * std.error)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_y_discrete(labels = c("tws_to_forest" = "TWS to Forest",
                              "tws_no_change" = "TWS No Change", 
                              "delta_recorder_effort" = "ΔRecorder Effort",
                              "log_recorder_effort" = "log Recorder Effort",
                              "lc_time_period2006-2012" = "2006-2012 Time Period",
                              "lc_time_period2012-2018" = "2012-2018 Time Period"),
                   limits = c("lc_time_period2012-2018",
                              "lc_time_period2006-2012", 
                              "log_recorder_effort",
                              "delta_recorder_effort",
                              "tws_no_change",
                              "tws_to_forest")) +
  labs(x = "Estimate ± 95% CI", y = NULL) +
  theme_classic()

# Display plot
figure7_b
```

# Bird Occurrences

```{r prepare bird data}
#| label: bird-data-prep
#| include: false

# Load data
load(here("data", "derived_data", 
          "birds_all_periods_turnover_all_land_cover_chanegs_15km.rda"))

# Keep only TWS -> Forest
birds_turnover_tws_forest_15km <- birds_all_periods_turnover_all_land_cover_chanegs_15km |>
  select(-c('2000-2006_Forest no change', '2000-2006_Forest to TWS',
            '2000-2006_Urban_no_change', '2000-2006_all_to_urban',
            '2006-2012_Forest no change', '2006-2012_Forest to TWS',
            '2006-2012_Urban_no_change', '2006-2012_all_to_urban',
            '2012-2018_Forest no change', '2012-2018_Forest to TWS',
            '2012-2018_Urban_no_change', '2012-2018_all_to_urban')) |>
  # determine which rows belong to which time period
  mutate(tws_no_change = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS no change`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_TWS no change`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_TWS no change`,
                                   TRUE ~ NA_real_),
         tws_to_forest = case_when(lc_time_period == "2000-2006" ~ `2000-2006_TWS to Forest`,
                                   lc_time_period == "2006-2012" ~ `2006-2012_TWS to Forest`,
                                   lc_time_period == "2012-2018" ~ `2012-2018_TWS to Forest`,
                                   TRUE ~ NA_real_)) |>
  # remove columns no longer required
  select(-`2000-2006_TWS no change`, -`2006-2012_TWS no change`, 
         -`2012-2018_TWS no change`,-`2000-2006_TWS to Forest`, 
         -`2006-2012_TWS to Forest`, -`2012-2018_TWS to Forest`)

# Remove rows that might have NA for x or y & categorise time periods
birds_turnover_tws_forest_15km_coords_time <- birds_turnover_tws_forest_15km |>
  filter(!is.na(x) & !is.na(y)) |>
  mutate(time_numeric = case_when(lc_time_period == "2000-2006" ~ 1,
                                  lc_time_period == "2006-2012" ~ 2,
                                  lc_time_period == "2012-2018" ~ 3),
         log_recorder_effort = log(recorder_effort))
```

## Summary Statistics

```{r bird summary statistics}
#| label: tbl-summary-stats-birds
#| echo: false
#| tbl-cap: "Summary Statistics of Jaccard Dissimilarity Index (JDI) by Time Period for Bird Records"

# Extract summary data
birds_summary_by_period <- birds_turnover_tws_forest_15km |>
  group_by(lc_time_period) |>
  summarize(n_observations = n(),
            mean_JDI = mean(JDI, na.rm = TRUE),
            sd_JDI = sd(JDI, na.rm = TRUE),
            mean_tws_to_forest = mean(tws_to_forest, na.rm = TRUE),
            sd_tws_to_forest = sd(tws_to_forest, na.rm = TRUE),
            mean_tws_no_change = mean(tws_no_change, na.rm = TRUE),
            sd_tws_no_change = sd(tws_no_change, na.rm = TRUE),
            mean_delta_recorder_effort = mean(delta_recorder_effort, na.rm = TRUE),
            mean_recorder_effort = mean(recorder_effort, na.rm = TRUE))

# Create table for period summary
birds_summary_by_period <- birds_summary_by_period |>
  kbl(digits = 3,
      align = c("l", rep("r", ncol(birds_summary_by_period)-1))) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE) |>
  row_spec(0, bold = TRUE, background = "#E6F0FF") |>
  add_header_above(c(" " = 1, 
                     "Sample" = 1,
                     "Turnover" = 2, 
                     "TWS to Forst" = 2,
                     "TWS No Change" = 2,
                     "Recorder Effort" = 2))

# Print the table
birds_summary_by_period
```

## Distribution of Values

```{r bird values}
#| label: fig-variables-distributions-birds
#| fig-cap: "Distribution of a) Jaccard Dissimilarity Index, b) Forest to TWS transition pixel count, c) Forest No Change pixel count, d) Recorder Effort e) Change in Recorder Effort in Subser with Only Bird Records"
#| echo: false

# Jaccard's Dissimilarity Index
bird_jdi_hist <- ggplot(birds_turnover_tws_forest_15km, aes(x = JDI)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Jaccard Dissimilarity Index",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# TWS -> Forest
bird_tws_to_forest_hist <- ggplot(birds_turnover_tws_forest_15km, aes(x = tws_to_forest)) +
  geom_histogram(bins = 30, fill = "brown", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "TWS to Forest Transition (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# TWS No Change
bird_tws_no_change_hist <- ggplot(birds_turnover_tws_forest_15km, aes(x = tws_no_change)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "TWS No Change (pixel count)",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Recorder effort
bird_recorder_effort_plot <- ggplot(birds_turnover_tws_forest_15km, aes(x = recorder_effort)) +
  geom_histogram(bins = 30, fill = "orange", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Delta recorder effort
bird_delta_recorder_effort_plot <- ggplot(birds_turnover_tws_forest_15km, aes(x = delta_recorder_effort)) +
  geom_histogram(bins = 30, fill = "coral", color = "white", alpha = 0.7) +
  facet_wrap(~ lc_time_period) +
  labs(x = "Delta Recorder Effort",
       y = "Count") +
  theme_classic() +
  theme(strip.background = element_rect(fill = "lightgrey"),
        strip.text = element_text(face = "bold"))

# Combine plots
bird_plots <- plot_grid(bird_jdi_hist, bird_tws_to_forest_hist, bird_tws_no_change_hist, bird_recorder_effort_plot, bird_delta_recorder_effort_plot, ncol = 2, labels = c("a)", "b)", "c)", "d)", "e)"))

# Print plots
bird_plots
```

## Correlation Matrix

```{r bird correlation matrix}
#| label: fig-variables-relationships-birds
#| echo: false
#| fig-cap: "Correlation matrix of variables used in the bird models"
#| fig-width: 6
#| fig-height: 4

# Correlation matrix for numeric variables
bird_numeric_vars <- birds_turnover_tws_forest_15km_coords_time |>
  select(JDI, tws_to_forest, tws_no_change, 
         intersection_size, union_size, 
         total_spp_before, total_spp_after,
         total_occ_before, total_occ_after,
         recorder_effort, delta_recorder_effort,
         log_recorder_effort)

# Calculate correlation matrix
bird_cor_matrix <- cor(bird_numeric_vars, use = "pairwise.complete.obs")

# Create correlation plot
bird_cor_plot <- corrplot(bird_cor_matrix, 
                     method = "circle", 
                     type = "upper", 
                     tl.col = "black", 
                     tl.srt = 45, 
                     addCoef.col = "black", 
                     number.cex = 0.7,
                     diag = FALSE)
```

## Modeling

### (Final) Trial 1: GLS with log recorder effort

```{r bird gls}
#| label: bird-gls-output
#| echo: false
#| message: false

# Load model
load(here("data", "models", "exploratory", "birds_TWSF_model1_gls.RData"))

# Extract summary
birds_TWSF_model1_gls_summary <- summary(birds_TWSF_model1_gls)

# Extract correlation structure parameters
print(birds_TWSF_model1_gls$modelStruct$corStruct)

# Get range parameter
birds_range_param <- coef(birds_TWSF_model1_gls$modelStruct$corStruct, unconstrained = FALSE)
```

```{r gls birds log}
#| label: fig-birds-gls-logged-values
#| echo: false
#| message: false
#| fig-width: 12
#| fig-height: 8
#| fig-cap: "Diagnostic plots for Bird GLS"

# Extract residuals
birds_residuals_gls <- residuals(birds_TWSF_model1_gls, type = "normalized")

# Get the data actually used in the model
birds_model_data <- getData(birds_TWSF_model1_gls)

# Basic residual plots
par(mfrow = c(2, 2))

# Residuals vs fitted
plot(fitted(birds_TWSF_model1_gls), birds_residuals_gls,
     xlab = "Fitted Values", ylab = "Normalized Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

# QQ plot
qqnorm(birds_residuals_gls, main = "Normal Q-Q Plot")
qqline(birds_residuals_gls, col = "red")

# Residuals vs predictors
plot(birds_model_data$tws_to_forest, birds_residuals_gls,
     xlab = "TWS to Forest", ylab = "Normalized Residuals",
     main = "Residuals vs TWS to Forest")
abline(h = 0, col = "red", lty = 2)

plot(birds_model_data$log_recorder_effort, birds_residuals_gls,
     xlab = "Recorder Effort", ylab = "Normalized Residuals",
     main = "Residuals vs Recorder Effort")
abline(h = 0, col = "red", lty = 2)
```

The diagnostic plots suggest that:

1.  Residuals vs Fitted:
    -   There is a downward slope and fan shape =\> suggests non-constant variance and possible misspecification of the model
2.  QQ Plot:
    -   Relatively linear but with deviations at both ends
3.  Residuals vs Forest to TWS
    -   Some banding at low values but not too strong
4.  Residuals vs Recorder Effort
    -   Some funneling =\> might be concerning

Despite some of the issues in the diagnostic plots, this is the model we will use to keep consistent across analyses.

```{r gls birds summary}
#| label: tbl-gls-birds-values-summary
#| echo: false
#| tbl-cap: "GLS Model Coefficients TWS → Forest Effects on Birds Turnover"

# Extract model summary
model_summary_birds <- summary(birds_TWSF_model1_gls)
coef_table_birds <- model_summary_birds$tTable

# Create coefficient table without confidence intervals initially
model_coefs_birds <- data.frame(Variable = rownames(coef_table_birds),
                          Estimate = round(coef_table_birds[, "Value"], 4),
                          `Std Error` = round(coef_table_birds[, "Std.Error"], 4),
                          `t value` = round(coef_table_birds[, "t-value"], 3),
                          `p value` = coef_table_birds[, "p-value"],
                          check.names = FALSE)

# Add significance stars
model_coefs_birds$Significance <- case_when(model_coefs_birds$`p value` < 0.001 ~ "***",
                                      model_coefs_birds$`p value` < 0.01 ~ "**", 
                                      model_coefs_birds$`p value` < 0.05 ~ "*",
                                      model_coefs_birds$`p value` < 0.1 ~ ".",
                                      TRUE ~ "")

# Display using kable
knitr::kable(model_coefs_birds, digits = c(0, 4, 4, 3, 6, 0),
             format.args = list(scientific = FALSE))
```

```{r gls birds model fit statistics}
#| label: tbl-gls-birds-model-fit
#| echo: false
#| tbl-cap: "GLS Birds Model Fit Statistics"


# Model information
cat("Model Specifications:\n")
cat("- Formula:", deparse(birds_TWSF_model1_gls$call$model), "\n")
cat("- Correlation Structure: Exponential spatial correlation\n") 
cat("- Method: REML\n")

# Try to get sample size safely
n_obs <- tryCatch({model_summary_birds$dims$N
}, error = function(e) {
  nobs(birds_TWSF_model1_gls)
})

cat("- Number of observations:", n_obs, "\n")

# Model fit statistics
fit_stats <- data.frame(Statistic = c("AIC", "BIC", "Log-likelihood", "Residual Std Error"),
                        Value = c(round(AIC(birds_TWSF_model1_gls), 2),
                                  round(BIC(birds_TWSF_model1_gls), 2), 
                                  round(as.numeric(logLik(birds_TWSF_model1_gls)), 2),
                                  round(model_summary_birds$sigma, 4)))
knitr::kable(fit_stats, col.names = c("Statistic", "Value"))
```

```{r gls bird model output plot}
#| label: fig-gls-bird-coefficient-plot
#| echo: false
#| message: false
#| warning: false
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Coefficient Estimates from Bird GLS"

# Get summary of the model
model_summary <- summary(birds_TWSF_model1_gls)

# Create dataframe of coeficients
birds_TWSF_model1_gls_coef_df <- data.frame(term = names(model_summary$tTable[, "Value"]),
                                                   estimate = model_summary$tTable[, "Value"],
                                                   std.error = model_summary$tTable[, "Std.Error"],
                                                   statistic = model_summary$tTable[, "t-value"],
                                                   p.value = model_summary$tTable[, "p-value"])

# Remove the intercept
birds_TWSF_model1_gls_coef_df_no_intercept <- birds_TWSF_model1_gls_coef_df[birds_TWSF_model1_gls_coef_df$term != "(Intercept)", ]

# Create coefficient plot
figure7_c <- ggplot(birds_TWSF_model1_gls_coef_df_no_intercept, aes(x = estimate, y = term)) +
  geom_point() +
  geom_errorbarh(aes(xmin = estimate - 1.96 * std.error,
                     xmax = estimate + 1.96 * std.error)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  scale_y_discrete(labels = c("tws_to_forest" = "TWS to Forest",
                              "tws_no_change" = "TWS No Change", 
                              "delta_recorder_effort" = "ΔRecorder Effort",
                              "log_recorder_effort" = "log Recorder Effort",
                              "lc_time_period2006-2012" = "2006-2012 Time Period",
                              "lc_time_period2012-2018" = "2012-2018 Time Period"),
                   limits = c("lc_time_period2012-2018",
                              "lc_time_period2006-2012", 
                              "log_recorder_effort",
                              "delta_recorder_effort",
                              "tws_no_change",
                              "tws_to_forest")) +
  labs(x = "Estimate ± 95% CI", y = NULL) +
  theme_classic()

# Display plot
figure7_c
```
